---
title: "Randomized Block Design Analysis of Wheat Yield"
author: "Joseph Mutua"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# 1. Objective

The objective of this study is to determine whether wheat yield differs:

1. Between geographical areas (treatments),
2. Between years (blocks),
3. And whether there is an interaction between area and year.

The experiment follows a Randomized Block Design (RBD) with replication, where:

- Treatments = 3 Areas
- Blocks = 3 Years
- Replications per cell = 2 plots
- Response = Yield (kg)

---

# 2. Statistical Model

The model used is:

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

where:

- $\mu$ = overall mean  
- $\alpha_i$ = area effect  
- $\beta_j$ = year effect  
- $(\alpha\beta)_{ij}$ = interaction  
- $\epsilon_{ijk} \sim N(0,\sigma^2)$

---

# 3. Hypotheses Tested (5% level)

### (i) Area effect
$H_0: \alpha_1 = \alpha_2 = \alpha_3$  
$H_1: \text{At least one } \alpha_i \text{ is different}$

### (ii) Year effect
$H_0: \beta_1 = \beta_2 = \beta_3$  
$H_1: \text{At least one } \beta_j \text{ is different}$

### (iii) Interaction
$H_0: (\alpha\beta)_{ij} = 0$  
$H_1: (\alpha\beta)_{ij} \neq 0$

---

# 4. Data Entry

```{r}
area  <- rep(c("A1", "A2", "A3"), each = 2, times = 6)
year  <- rep(c("Y1", "Y2", "Y3"), each = 12)
repno <- rep(1:2, times = 18)

yield <- c(
14.20,17.45,13.6,19.22,16.96,16.3,
14.40,16.00,16.28,25.4,16.62,14.3,
14.98,15.9,17.02,14.3,12.16,14.34,
15.80,17.58,13.96,13.1,16.89,13.46,
14.34,12.14,12.0,14.74,17.83,16.22,
11.96,15.24,14.5,12.86,18.03,15.12)

wheat <- data.frame(area, year, repno, yield)

head(wheat)
```



#  5. Fit the RBD Model
```{r}
model <- aov(yield ~ area * year, data = wheat)
summary(model)
```


#  6. Results and Interpretation
From the ANOVA table: 
Area: $p=0.7161 > 0.05$ (Not significant)


Year: $p=0.0547 > 0.05$ (Not significant)


Interaction (Area × Year): $p=0.0413 < 0.05$ (Significant)

Since the interaction is significant, the effect of the geographical area on yield depends on which year is being observed.


# 7. ConclusionAt the 5% level of significance:

We accept $H_0$ for the Area effect; the geographical areas are not significantly different on average.


We accept $H_0$ for the Year effect; the blocks (years) are not significantly different on average.



We reject $H_0$ for the Interaction; there is a significant interaction between area and year.




Final Note: Because the interaction is significant, we should focus on the specific combinations of Area and Year rather than looking at the average performance of an Area across all years.
